A framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall

### 2019-06-19 paper/main.tex
updated out-of-date bibtex keys, recompiled pdf

How might one go about formally characterizing the shape of an experience, or whether that shape has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

@@

Our work casts remembering as reproducing (behaviorally and neurally) the topic trajectory, or shape, of the original experience.  This view draws inspiration from prior work aimed at elucidating the neural and behavioral underpinnings of how we process dynamic naturalistic experiences and remember them later.  One approach to identifying neural responses to naturalistic stimuli (including experiences) entails building a model of the stimulus and searching for brain regions whose responses are consistent with the model.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have extended this approach with a clever twist.  Rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time.  By contrast, in our present work we used topic models and HMMs to construct an explicit stimulus model (i.e., the topic trajectory of the video).  When we searched for brain structures whose responses are consistent with the video's topic trajectory, we identified a network of structures that overlapped strongly with the ``long temporal receptive window'' network reported by the Hasson group~\citep[e.g., compare our Fig.~\ref{fig:brainz}A with the map of long temporal receptive window voxels in][]{LernEtal11}.  This provides support for the notion that part of the long temporal receptive window network may be maintaining an explicit model of the stimulus dynamics.  When we performed a similar analysis after swapping out the video's topic trajectory with the recall topic trajectories of each individual participant, this allowed us to identify brain regions whose responses (as the participants viewed the video) reflected how the video trajectory would be transformed in memory (as reflected by the recall topic trajectories).  The analysis revealed that the rMTL and vmPFC may play a role in this person-specific transformation from experience into memory.  The role of the MTL in episodic memory encoding has been well-reported~\citep[e.g., ][]{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.  Prior work has also implicated the medial prefrontal cortex in representing ``schema'' knowledge~\citep[i.e., general knowledge about the format of an ongoing experience given prior similar experiences; ][]{KestEtal12, SchlPres15, GilbMarl17, SpalEtal18}.  Integrating across our study and this prior work, one interpretation is that the person-specific transformations mediated (or represented) by the rMTL and vmPFC may reflect schema knowledge being leveraged, formed, or updated, incorporating ongoing experience into previously acquired knowledge.

>>>>

How might one go about formally characterizing the shape of an experience, or whether that shape has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

@@

Our work casts remembering as reproducing (behaviorally and neurally) the topic trajectory, or shape, of the original experience.  This view draws inspiration from prior work aimed at elucidating the neural and behavioral underpinnings of how we process dynamic naturalistic experiences and remember them later.  One approach to identifying neural responses to naturalistic stimuli (including experiences) entails building a model of the stimulus and searching for brain regions whose responses are consistent with the model.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have extended this approach with a clever twist.  Rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time.  By contrast, in our present work we used topic models and HMMs to construct an explicit stimulus model (i.e., the topic trajectory of the video).  When we searched for brain structures whose responses are consistent with the video's topic trajectory, we identified a network of structures that overlapped strongly with the ``long temporal receptive window'' network reported by the Hasson group~\citep[e.g., compare our Fig.~\ref{fig:brainz}A with the map of long temporal receptive window voxels in][]{LernEtal11}.  This provides support for the notion that part of the long temporal receptive window network may be maintaining an explicit model of the stimulus dynamics.  When we performed a similar analysis after swapping out the video's topic trajectory with the recall topic trajectories of each individual participant, this allowed us to identify brain regions whose responses (as the participants viewed the video) reflected how the video trajectory would be transformed in memory (as reflected by the recall topic trajectories).  The analysis revealed that the rMTL and vmPFC may play a role in this person-specific transformation from experience into memory.  The role of the MTL in episodic memory encoding has been well-reported~\citep[e.g., ][]{PallWagn02, DavaEtAl03, RangEtal04a, Dava06}.  Prior work has also implicated the medial prefrontal cortex in representing ``schema'' knowledge~\citep[i.e., general knowledge about the format of an ongoing experience given prior similar experiences; ][]{KestEtal12, SchlPres15, GilbMarl17, SpalEtal18}.  Integrating across our study and this prior work, one interpretation is that the person-specific transformations mediated (or represented) by the rMTL and vmPFC may reflect schema knowledge being leveraged, formed, or updated, incorporating ongoing experience into previously acquired knowledge.
